Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2017-03-28 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
fixed % and abstract

I would also like to thank Roger Nokes for his development and continued assistance with Streams and to the New Zealand Fire Commission for their financial support to the University of Canterbury

Last but not least, I thank my wife, Liz, for her patience and support of my pursuit of higher education throughout our 38 years of marriage.  I would also like to thank my children; Catherine, Stephanie, Michael, Christina, and Veronica for ``buying me time'' to work on my studies.}

\abstract{
Unwanted fires result in loss of life and property.  These fires can also create an adverse economic impact on a community.  The investigation of fires provides a means to identify the cause of the fire in order to develop a knowledge base that could enable the elimination of that cause and thus reduce the losses from unwanted fires. Questions about the lack of science in the practice of fire investigation have been raised during the review of several arson homicide cases and a forensics science review by the U.S. National Academy of Sciences.  Specifically the National Academy of Sciences indicated that ``... research is needed on the natural variability of burn patterns...''
This study addresses that need in two ways:
\begin{enumerate}
\item Examining the repeatability of several small fire sources and the fire patterns that were generated by those fires
\item Examining the capability of numerical models to simulate the fires and the resulting fire patterns based on input data collected from engineering reference sources, and bench-scale and full-scale fire experiments conducted in this study.
\end{enumerate}
This manuscript highlights the many uncertainties involved in what appear to be simple fire experiments.  Uncertainties related to fuel type, measurement methods, and analysis techniques were examined.  The objective of the research was to examine the repeatability of pre-flashover fire patterns generated from exposure to short duration (300 s maximum) well characterized fires.  Three different fuels were used: natural gas, gasoline, and polyurethane foam.  Each fuel had a similar top surface area.  The heat release rate data showed that the variability between replicates was greater for more complex fuels.  The variation in peak heat release rate with the natural gas was similar to the expanded uncertainty of the measurement system, 11~\%.  However, the variation in the peak heat release rates of the gasoline and the polyurethane foam increased due to increased uncertainties in the burning behavior of the fuels.

The variability in the heat release rate of the fires resulted in higher levels of variability of the replicate fire patterns. The maximum fire pattern heights generated from the natural gas and gasoline fires were shown to have uncertainties of 18~\% or less based on a Type A statistical analysis with 95~\% confidence limits.  The comparison of the fire pattern heights and the mean flame height demonstrated that the ``steady state'' natural gas fires exhibited the highest level of agreement and polyurethane foam fueled fire exhibited the least agreement, with the gasoline fueled fires in between.

A range of methods used to simulate characteristics of fires were applied using data from the source fire experiments as input.  The algebra based predictions for heat flux to a target under predicted the measured values.  The gasoline fueled fires best agreement of 22~\% or less for the heat flux predictions.  The polyurethane foam results exhibited the worst agreement with difference of at least 40~\% between the measured and calculated values.  Improved understanding of the radiative fraction from different fuels would improve the capability of the predictions.

The best agreement between measured flame height, the measured height of the fire pattern, and flame height predictions occurred with the gasoline fueled fires.  Given the overlap of the expanded uncertainty and predicted range of values, the measured and predicted heights would be considered similar.  The gasoline fueled fires released the highest amount of energy and has a higher radiative fraction than the natural gas and polyurethane foam fueled fires.

The agreement between the computational fluid dynamics predicted values and the measurements from these experiments demonstrated examples of agreement that were within the measurement uncertainty estimates as well as examples with poor agreement.  The differences were driven in some cases by limitations within the combustion sub-model close to the burner surface, as the distance between the burner surface increased the predicted values tended to converge with the measured values.  The wall located adjacent to the burn had a significant impact on the flow field in and around the flame and plume region.  The horizontal flame movements observed during the experiments was not simulated to the same degree by the model.  Given the areas around the flame have steep thermal gradients small differences in position can result in large differences in temperature and heat flux.

The models are useful tools for gaining insight into fire behavior but in many cases many require specific experimental data for input or for validation. The availability   of well characterized input data is limited and data for newer materials is unavailable. Additional research is needed for modeling data development protocols.  Improved data repositories are also needed. Further research is needed to understand the appropriate resolution for implementing the results based on fuel type, fuel geometry, and ventilation.  Most importantly, there is a critical need to extend this research to better understand the capabilities and limitations of predictive methods using fuels and scenarios that fire investigators are likely to encounter during a fire investigation.  
}

@@

List of Symbols
\begin{tabbing}
\hspace{1.5in} \= \\
$c_{p}$ \> specific heat \\
$k$ \> thermal conductivity \\
$\chi_r$ \> radiative fraction \\
$\dot{q}''$ \> heat flux per unit area kW/m$^2$ \\
$\dot{Q}$ \> heat release rate (kW) \\ 
$R$ \> distance from the point source to the target (m) \\
$\rho$ \> density \\
\end{tabbing}

@@

Madrzykowski, D. and Fleischmann, C., "Fire Pattern Repeatability: A Laboratory Study on Gypsum Wallboard." In Proceedings ISFI 2010: International Symposium on Fire Investigation Science and Technology, Hyattsville, MD, USA. Sarasota, FL, USA 2010. pp 381-392.

@@

Madrzykowski, D. and Fleischmann, C., "Flame Heights and Heat Transfer from Gaseous, Liquid and Solid Fuels for Burn Pattern Repeatability Research." In INTERFLAM 2010: 12th International Conference on Fire Science and Engineering, Nottingham, UK. London, UK, 2010. pp 1930-1942.

@@

According to the NFPA, there were approximately 282,600 intentionally set fires each year during the period of 2007 through 2011 in the United States. The average annual loss totals for that period were approximately 420 civilian deaths and 1,360 civilian injuries. More than \$1.3 billion dollars (U.S.) was lost per year due to direct property damage. Structure fires represent less than 20~\% of intentionally set fires.  However, fires involving structures account for more than 84~\% of the civilian deaths, injuries and property loss from all intentional fires.  Approximately two-thirds of the intentional structure fires occurred in occupied and operating residential occupancies.  As a result, 95~\% of the civilian deaths and 86~\% of the civilian injuries which are attributed to intentional fires occurred in residential occupancies~\cite{Campbell:2014}.

A subset of the intentional fires are considered arson.  The U.S. Federal Bureau of Investigation (FBI) Uniform Crime Reporting (UCR) Program defines arson as any willful or malicious burning or attempting to burn, with or without intent to defraud, a dwelling house, public building, motor vehicle or aircraft, personal property of another, etc.~\cite{Crime:2010}.  The national rate of arson fire cases cleared by arrest or exceptional means is less than 20~\%.  Of those cases approximately 46~\% of the suspects were prosecuted~\cite{Campbell:2014}.  Hence less than 10~\% of the arson cases resulted in successful prosecution.

@@

Flexible polyurethane foam was chosen as the solid fuel. According to the Polyurethane Foam Association, flexible polyurethane foam is the most widely used cushioning material in upholstered furnishing and mattresses. More than 1.2 billion pounds are produced in the U.S. every year~\cite{Foam,Polyurethane_Foam}. In studies by Ahrens and Evarts of the NFPA, upholstered furniture, and mattresses and bedding were the first items ignited in 33~\% of the fatal fires in the U.S. based on averaged data from 2005 through 2009~\cite{Ahrens:2011,Evarts:2011}.

@@

The NIST Large Fire Research Laboratory's 3~m by 3~m oxygen depletion calorimeter had an estimated peak heat release rate capacity of approximately 700~kW.  Details on the operation and uncertainty in measurements associated with NIST’s 6~m by 6~m oxygen consumption calorimeter were documented by Bryant et al.~\cite{Bryant:2004}. Many of the same instruments were shared between the two calorimeters and calibration burns were conducted during each experimental series.  As a result, the expanded uncertainty of this device has been estimated as 11~\% on the measured heat release rate.

@@

Gas temperatures were measured with exposed-bead, Chromel-Alumel (type K) thermocouples, with a 1.0~mm nominal diameter.  Starting from the exposed bead, the thermocouple wire was sheathed in a 3.2~mm diameter inconel shield, 0.76~m in length. The thermocouple was ungrounded. The standard uncertainty in temperature of the thermocouple wire itself is 2.2~$^\circ$C at 277~$^\circ$C and increases to 9.5~$^\circ$C at 871~$^\circ$C as determined by the wire manufacturer.  The response time constant is approximately 0.2~seconds~\cite{Omega}.  The variation of the temperature in the environment surrounding the thermocouple was known to be much greater than that of the wire uncertainty~\cite{Blevins:1999,Pitts:2001}. Small diameter thermocouples were used to limit the impact of radiative heating and cooling.  The estimated combined expanded uncertainty for temperature in these experiments was 15~\%.

@@

The heat flux was measured with water cooled, Schmidt-Boelter total heat flux gauges. The sensing surface of the heat flux gauges were located 0.95~m horizontally from the center of the burner/fuel and 0.50~m above the top of the burner/pan lip/fuel surface.  The manufacturer reports a 3~\% calibration expanded uncertainty for these devices~\cite{Medtherm}. Results from an international study on total heat flux gauge calibration and response demonstrated that the uncertainty of a Schmidt-Boelter gauge was typically 8~\%~\cite{Pitts:2006}.  The gauges were mounted to uprights of the steel instrumentation support frame on opposite sides of the fire.

@@

The fires were centered under the hood of the calorimeter on a non-combustible platform measuring 0.91~m x 1.2~m x 12~mm thick supported by the load cell.  The load cell had a resolution of 1.0~g~\cite{Mettler}. The expanded uncertainty was estimated to be 5~\%.  The load cell was checked between experiments with calibrated weights during the zero and span checks of the calorimeter.

@@

A schematic drawing of the instrumentation arrangement is shown in Figure~\ref{Drawing_Fire_Flamer}.  The thermocouples were mounted on a steel frame.  The frame enabled the thermocouples to be installed at 10~cm intervals from 10~cm to 120~cm above the burner/pan lip/fuel surface. At each 10~cm interval, there were five thermocouples: one centered above the burner/fuel surface and one centered above each edge of the burner/fuel surface, for a total of 70 thermocouples.  A plumb bob and steel tape measure, with a resolution of $\pm$~0.5~mm, were used to position and align the thermocouples.   The expanded uncertainty for the location of each thermocouple tip is estimated to be $\pm$~2.5~\%.  The same applies to the location of the heat flux sensor surfaces.

@@

On the averaged heat release rate curve for each fuel, the shaded area surrounding the average represents the 95~\% confidence level (2$\sigma$), based on Type A evaluation of the uncertainty is provided~\cite{Taylor:1994}.  A Type A evaluation of the uncertainty is any statistical analysis of the data from a series of observations.  In this case the evaluation assumed a normal distribution of the data such that doubling the calculated standard deviation (2$\sigma$) would yield a confidence level of approximately 95~\%.

The natural gas burner provided the most repeatable results, with an average steady state heat release rate of 80~kW~$\pm$~9~kW with an average total heat release of 23.3~MJ~$\pm$~1.6~MJ over a 300~s period. The uncertainty of the steady state heat release rate measurement is approximately~$\pm$~11~\% It should be noted that 11~\% is also the (2$\sigma$) uncertainty estimate of the calorimeter. The gas burner and on/off control valve allowed the heat release rate to be similar in form to a ``square-wave''.

The gasoline pan fire was ignited and then burned until the fuel was consumed.  Therefore the heat release rate curves had more of a ``bell shape''  as opposed to the flat steady- state peak heat release rate region of the natural gas burner.  Therefore the  gasoline average peak heat release rate was based on an average over a 30 s interval bounding the peak heat release rate of each of the 18 experiments.  This method provided an average peak heat release rate of 80~kW~$\pm$~24~kW or 30~\%.  The total heat release of the gasoline over the 300~s period after ignition was 13.6~MJ~$\pm$~0.6~MJ.

The block of polyurethane foam demonstrated the largest variability of the three fuels in terms of peak heat release rate and time to peak heat release. Similar to the gasoline that began with a fixed quantity of fuel which was ignited and burned to completion, the polyurethane foam also generated ``bell shaped'' heat release rate curve.  Since the duration of the heat release rate peaks were less than those of the gasoline, the peak heat release rate averaging was performed over a 15~s interval bounding the peak heat release rate from each of the ten experiments.  This process yielded an average peak heat release rate of 45~kW~$\pm$~21~kW or 47~\%.  The total heat release over the 300~s period after ignition was 4.1~MJ~$\pm$~0.2~MJ.

The Figures~\ref{HRR}-\ref{HRR2} have a shaded area that shows the 95~\% confidence level.  The more narrow the shaded area, the better the repeatability of the fire.  This provides a quick means to assess the repeatability of a given source fire.  Using this measure the natural gas burner fires were the most repeatable, followed by the gasoline pan fires and then the polyurethane fires.

@@

In engineering terms, the ``flame height'' is typically discussed as the median flame height.  A position above the top of the fuel surface where the flame is observed to be present at least 50~\% of the time.  As a starting point, the mean flame height for each fuel was estimated from a measurement of the uppermost continuous flame tip from the still photographic images.

@@

Another means of quantifying field data with Streams is the use of contour plots.  Examples of the time averaged contour plots are shown in Figures~\ref{Gas_and_Gasoline_Contours} and \ref{Foam_Contours} for the natural gas, gasoline, and polyurethane fueled fires.  Each contour line represents the percentage of time that the flame was at a given position. The region under examination, as shown as a black bounding rectangle, is 1400~mm high and 800~mm in width.  The natural gas fire was sampled for 29~s, resulting in 870~images. The area bounded by the contour with the value 1 represents the area that the flame was visible 100~\% of the time sampled. This can be considered the persistent flame region as defined by McCaffrey~\cite{McCaffrey:1979}. The remaining contours are in the intermittent flame region, with the 0.5 contour representing the mean flame height. Outside of the final contour, value 0.0, no visible flame was recorded; hence this contour marks the definitive start of the buoyant plume region.

@@

Figure~\ref{Foam_Contours} shows two different contours for the polyurethane foam fueled experiment.  The results from polyurethane fueled fire characterization experiments have an increased amount of variability due to the nature of the ignition source and the fuel itself.  The polyurethane foam was readily ignited with a small pilot flame similar to that used to ignite the natural gas and the gasoline.  By comparison, a small flame ignited the gases flowing out of the top of the natural gas burner and the flames from the natural gas cover the entire burner surface within approximately one second.  A similar ignition sequence occurs with the gasoline pan fires. The polyurethane foam has a more complicated flame spread sequence that resulted in significant changes to the peak heat release rate and the time to peak heat release.  Once the polyurethane foam was ignited, it took approximately 60~s for the flame to spread across the top surface of the fuel.  As the flames spread horizontally across the fuel surface, the fire also burns down into the fuel.  In some cases the fire burned completely through the fuel in the center of the foam sample, resulting in a hollow flame at the time of the peak heat release rate.  An example of a contour plot, exhibiting the hollow or bifurcated flames is shown at right in Figure~\ref{Foam_Contours}. Refer back to Figure~\ref{flame_height_comp_FHPUF_1} to see photographic views of the bifurcated flames.  It seems that the movement of the flames as they flowed away from the fuel surface blurred out the bifurcation over time with the exception of the lower elevation areas that presented at the 90 and 100~\% contours.  Other methods of ignition for the polyurethane foam were considered in order to have a more uniform ignition of the top surface of the foam.  Each method would have resulted in a larger release of energy from the ignition source or a starter fuel.   Based on a few scoping experiments these methods did not improve the repeatability.

Based on the field data developed using Streams average flame heights for each fuel were determined.  In Table~\ref{tab:Average_Flame_Heights}, each of the values represents the average flame heights, given in mm with 95~\% confidence limits based on a Type A statistical analysis of the data, for each of the source fuels shown as the percentage of time the flame was visible at that given height.  For the natural gas fires, the averages are from ten 30~s segments, while the fire was monitored at approximately 80~kW.  In some cases more than one 30~s segment was sampled per heat release rate experiment.  Sixteen 30~s video segments, which included the peak heat release rate, were successfully analyzed from the seventeen gasoline heat release rate experiments.  One of the videos was unusable for analysis due to reflected light which could not be edited out of the images.  For the same reason, the polyurethane foam flame height averages are the result from eight of the ten heat release rate experiments.

\begin{table}
\centering
\begin{tabular}{|c|c|c|c|}
\hline
Visible         &   \multicolumn{3}{|c|}{Visible Flame Heights (mm)} \\ \cline{2-4}
Fraction (\%)   &       Natural Gas	            &   Gasoline	            & Polyurethane Foam \\ \hline \hline
10              &       980 $\pm$ 155 (16~\%)   &	930 $\pm$ 130 (14~\%) 	& 650 $\pm$ 220 (34~\%)   \\
20              &   	890 $\pm$ 125 (14~\%)   &	850 $\pm$ 120 (14~\%) 	& 590 $\pm$ 190 (32~\%)   \\
30              &   	820 $\pm$ 125 (15~\%)   &	790 $\pm$ 110 (14~\%) 	& 540 $\pm$ 180 (33~\%)   \\
40              &   	760 $\pm$ 105 (14~\%)   &	740 $\pm$ 95 (13~\%) 	& 500 $\pm$ 165 (33~\%)   \\
50              &   	705 $\pm$ 106 (15~\%)   &	695 $\pm$ 95 (14~\%) 	& 470 $\pm$ 160 (34~\%)   \\
60              &   	650 $\pm$ 100 (15~\%)   &	655 $\pm$ 85 (13~\%) 	& 435 $\pm$ 135 (31~\%)   \\
70              &   	590 $\pm$ 90 (15~\%)    &	610 $\pm$ 90 (15~\%) 	& 400 $\pm$ 125 (31~\%)   \\
80              &   	530 $\pm$ 85 (16~\%)    &	555 $\pm$ 90 (16~\%) 	& 360 $\pm$ 110 (31~\%)   \\
90              &   	440 $\pm$ 75 (17~\%)    &	490 $\pm$ 95 (19~\%) 	& 315 $\pm$ 110 (35~\%)   \\
100	            &       220 $\pm$ 84 (38~\%)    &	350 $\pm$ 110 (32~\%) 	& 210 $\pm$ 125 (60~\%)   \\
\hline
\end{tabular}
 \caption[Average flame heights for three fuels]{Average flame heights with 95~\% confidence limits, for each of the source fuels shown as the percentage of time the flame was visible at that given height.}
 \label{tab:Average_Flame_Heights}
\end{table}

>>>>

I would also like to thank Dr. Roger Nokes for his development and continued assistance with Streams, and to the New Zealand Fire Commission for their financial support to the University of Canterbury.

Last but not least, I thank my wife, Liz, for her patience and support of my pursuit of higher education throughout our 38 years of marriage.  I would also like to thank my children; Catherine, Stephanie, Michael, Christina, and Veronica, for buying me time to work on my studies.}

\abstract{
Unwanted fires result in loss of life and property.  These fires can also create an adverse economic impact on a community.  The investigation of fires provides a means to identify the cause of the fire in order to develop a knowledge base that could enable the elimination of that cause and thus reduce the losses from unwanted fires. Questions about the lack of science in the practice of fire investigation have been raised during the review of several arson homicide cases and a forensics science review by the U.S. National Academy of Sciences.  Specifically, the National Academy of Sciences indicated that ``... research is needed on the natural variability of burn patterns...''
This study addresses that need in two ways:
\begin{enumerate}
\item Examining the repeatability of several small fire sources and the fire patterns that were generated by those fires.
\item Examining the capability of numerical models to simulate the fires and the resulting fire patterns based on input data collected from engineering reference sources, bench-scale experiments, and full-scale fire experiments.
\end{enumerate}

This manuscript highlights the many uncertainties involved in what appear to be simple fire experiments.  Uncertainties related to fuel type, measurement methods, and analysis techniques were examined.  The objective of the research was to examine the repeatability of pre-flashover fire patterns generated from exposure to short duration (300 s maximum), well characterized fires.  Three different fuels were used: natural gas, gasoline, and polyurethane foam.  Each fuel had a similar top surface area.  The heat release rate data showed that the variability between replicates was greater for more complex fuels.  The variation in peak heat release rate with the natural gas was similar to the expanded uncertainty of the measurement system, 11\%.  However, the variation in the peak heat release rates of the gasoline and the polyurethane foam increased due to increased uncertainties in the burning behavior of the fuels.

The variability in the heat release rate of the fires resulted in higher levels of variability of the replicate fire patterns. The maximum fire pattern heights generated from the natural gas and gasoline fires were shown to have uncertainties of 18\% or less based on a Type A statistical analysis with 95\% confidence limits.  The comparison of the fire pattern heights and the mean flame height demonstrated that the steady state natural gas fires exhibited the highest level of agreement and polyurethane foam fueled fire exhibited the least agreement, with the gasoline fueled fires in between.

A range of methods used to simulate characteristics of fires were applied using data from the source fire experiments as input.  The algebra-based predictions for heat flux to a target under-predicted the measured values.  The gasoline fueled fires exhibited best agreement of 22\% or less for the heat flux predictions.  The polyurethane foam results exhibited the worst agreement with a difference of at least 40\% between the measured and calculated values.  Improved understanding of the radiative fraction from different fuels would improve the capability of the predictions.

The best agreement between measured flame height, the measured height of the fire pattern, and flame height predictions occurred with the gasoline fueled fires.  Given the overlap of the expanded uncertainty and predicted range of values, the measured and predicted heights would be considered similar.  The gasoline fueled fires released the highest amount of energy and had a higher radiative fraction than the natural gas and polyurethane foam fueled fires.

The agreement between the computational fluid dynamics predicted values and the measurements from these experiments demonstrated examples of agreement that were within the measurement uncertainty estimates as well as examples with poor agreement.  The differences were driven in some cases by limitations within the combustion sub-model close to the burner surface; as the distance between the burner surface increased, the predicted values tended to converge with the measured values.  The wall located adjacent to the burn had a significant impact on the flow field in and around the flame and plume region.  The horizontal flame movements observed during the experiments was not simulated to the same degree by the model.  Given the areas around the flame have steep thermal gradients, small differences in position can result in large differences in temperature and heat flux.

The models are useful tools for gaining insight into fire behavior but in many cases many require specific experimental data for input or for validation. The availability of well characterized input data is limited, and data for newer materials is unavailable. Additional research is needed for modeling data development protocols.  Improved data repositories are also needed. Further research is needed to understand the appropriate resolution for implementing the results based on fuel type, fuel geometry, and ventilation.  Most importantly, there is a critical need to extend this research to better understand the capabilities and limitations of predictive methods using fuels and scenarios that fire investigators are likely to encounter during a fire investigation.  
}

@@

List of Symbols
\begin{tabbing}
\hspace{1.5in} \= \\
$c_{p}$ \> specific heat \\
$k$ \> thermal conductivity \\
$\chi_r$ \> radiative fraction \\
$\dot{q}''$ \> heat flux per unit area (kW/m$^2$) \\
$\dot{Q}$ \> heat release rate (kW) \\ 
$R$ \> distance from the point source to the target (m) \\
$\rho$ \> density \\
\end{tabbing}

@@

Madrzykowski, D. and Fleischmann, C., `Fire Pattern Repeatability: A Laboratory Study on Gypsum Wallboard.' In Proceedings ISFI 2010: International Symposium on Fire Investigation Science and Technology, Hyattsville, MD, USA. Sarasota, FL, USA 2010. pp 381-392.

@@

Madrzykowski, D. and Fleischmann, C., `Flame Heights and Heat Transfer from Gaseous, Liquid, and Solid Fuels for Burn Pattern Repeatability Research.' In INTERFLAM 2010: 12th International Conference on Fire Science and Engineering, Nottingham, UK. London, UK, 2010. pp 1930-1942.

@@

According to the NFPA, there were approximately 282,600 intentionally set fires each year during the period of 2007 through 2011 in the United States. The average annual loss totals for that period were approximately 420 civilian deaths and 1,360 civilian injuries. More than \$1.3 billion dollars (U.S.) was lost per year due to direct property damage. Structure fires represent less than 20\% of intentionally set fires.  However, fires involving structures account for more than 84\% of the civilian deaths, injuries and property loss from all intentional fires.  Approximately two-thirds of the intentional structure fires occurred in occupied and operating residential occupancies.  As a result, 95\% of the civilian deaths and 86\% of the civilian injuries which are attributed to intentional fires occurred in residential occupancies~\cite{Campbell:2014}.

A subset of the intentional fires are considered arson.  The U.S. Federal Bureau of Investigation (FBI) Uniform Crime Reporting (UCR) Program defines arson as any willful or malicious burning or attempting to burn, with or without intent to defraud, a dwelling house, public building, motor vehicle or aircraft, personal property of another, etc.~\cite{Crime:2010}.  The national rate of arson fire cases cleared by arrest or exceptional means is less than 20\%.  Of those cases approximately 46\% of the suspects were prosecuted~\cite{Campbell:2014}.  Hence less than 10\% of the arson cases resulted in successful prosecution.

@@

Flexible polyurethane foam was chosen as the solid fuel. According to the Polyurethane Foam Association, flexible polyurethane foam is the most widely used cushioning material in upholstered furnishing and mattresses. More than 1.2 billion pounds are produced in the U.S. every year~\cite{Foam,Polyurethane_Foam}. In studies by Ahrens and Evarts of the NFPA, upholstered furniture, and mattresses and bedding were the first items ignited in 33\% of the fatal fires in the U.S. based on averaged data from 2005 through 2009~\cite{Ahrens:2011,Evarts:2011}.

@@

The NIST Large Fire Research Laboratory's 3~m by 3~m oxygen depletion calorimeter had an estimated peak heat release rate capacity of approximately 700~kW.  Details on the operation and uncertainty in measurements associated with NIST’s 6~m by 6~m oxygen consumption calorimeter were documented by Bryant et al.~\cite{Bryant:2004}. Many of the same instruments were shared between the two calorimeters and calibration burns were conducted during each experimental series.  As a result, the expanded uncertainty of this device has been estimated as 11\% on the measured heat release rate.

@@

Gas temperatures were measured with exposed-bead, Chromel-Alumel (type K) thermocouples, with a 1.0~mm nominal diameter.  Starting from the exposed bead, the thermocouple wire was sheathed in a 3.2~mm diameter inconel shield, 0.76~m in length. The thermocouple was ungrounded. The standard uncertainty in temperature of the thermocouple wire itself is 2.2~$^\circ$C at 277~$^\circ$C and increases to 9.5~$^\circ$C at 871~$^\circ$C as determined by the wire manufacturer.  The response time constant is approximately 0.2~seconds~\cite{Omega}.  The variation of the temperature in the environment surrounding the thermocouple was known to be much greater than that of the wire uncertainty~\cite{Blevins:1999,Pitts:2001}. Small diameter thermocouples were used to limit the impact of radiative heating and cooling.  The estimated combined expanded uncertainty for temperature in these experiments was 15\%.

@@

The heat flux was measured with water cooled, Schmidt-Boelter total heat flux gauges. The sensing surface of the heat flux gauges were located 0.95~m horizontally from the center of the burner/fuel and 0.50~m above the top of the burner/pan lip/fuel surface.  The manufacturer reports a 3\% calibration expanded uncertainty for these devices~\cite{Medtherm}. Results from an international study on total heat flux gauge calibration and response demonstrated that the uncertainty of a Schmidt-Boelter gauge was typically 8\%~\cite{Pitts:2006}.  The gauges were mounted to uprights of the steel instrumentation support frame on opposite sides of the fire.

@@

The fires were centered under the hood of the calorimeter on a non-combustible platform measuring 0.91~m x 1.2~m x 12~mm thick supported by the load cell.  The load cell had a resolution of 1.0~g~\cite{Mettler}. The expanded uncertainty was estimated to be 5\%.  The load cell was checked between experiments with calibrated weights during the zero and span checks of the calorimeter.

@@

A schematic drawing of the instrumentation arrangement is shown in Figure~\ref{Drawing_Fire_Flamer}.  The thermocouples were mounted on a steel frame.  The frame enabled the thermocouples to be installed at 10~cm intervals from 10~cm to 120~cm above the burner/pan lip/fuel surface. At each 10~cm interval, there were five thermocouples: one centered above the burner/fuel surface and one centered above each edge of the burner/fuel surface, for a total of 70 thermocouples.  A plumb bob and steel tape measure, with a resolution of $\pm$~0.5~mm, were used to position and align the thermocouples.   The expanded uncertainty for the location of each thermocouple tip is estimated to be $\pm$~2.5\%.  The same applies to the location of the heat flux sensor surfaces.

@@

On the averaged heat release rate curve for each fuel, the shaded area surrounding the average represents the 95\% confidence level (2$\sigma$), based on Type A evaluation of the uncertainty is provided~\cite{Taylor:1994}.  A Type A evaluation of the uncertainty is any statistical analysis of the data from a series of observations.  In this case the evaluation assumed a normal distribution of the data such that doubling the calculated standard deviation (2$\sigma$) would yield a confidence level of approximately 95\%.

The natural gas burner provided the most repeatable results, with an average steady state heat release rate of 80~kW~$\pm$~9~kW with an average total heat release of 23.3~MJ~$\pm$~1.6~MJ over a 300~s period. The uncertainty of the steady state heat release rate measurement is approximately~$\pm$~11\% It should be noted that 11\% is also the (2$\sigma$) uncertainty estimate of the calorimeter. The gas burner and on/off control valve allowed the heat release rate to be similar in form to a ``square-wave''.

The gasoline pan fire was ignited and then burned until the fuel was consumed.  Therefore the heat release rate curves had more of a ``bell shape''  as opposed to the flat steady- state peak heat release rate region of the natural gas burner.  Therefore the  gasoline average peak heat release rate was based on an average over a 30 s interval bounding the peak heat release rate of each of the 18 experiments.  This method provided an average peak heat release rate of 80~kW~$\pm$~24~kW or 30\%.  The total heat release of the gasoline over the 300~s period after ignition was 13.6~MJ~$\pm$~0.6~MJ.

The block of polyurethane foam demonstrated the largest variability of the three fuels in terms of peak heat release rate and time to peak heat release. Similar to the gasoline that began with a fixed quantity of fuel which was ignited and burned to completion, the polyurethane foam also generated ``bell shaped'' heat release rate curve.  Since the duration of the heat release rate peaks were less than those of the gasoline, the peak heat release rate averaging was performed over a 15~s interval bounding the peak heat release rate from each of the ten experiments.  This process yielded an average peak heat release rate of 45~kW~$\pm$~21~kW or 47\%.  The total heat release over the 300~s period after ignition was 4.1~MJ~$\pm$~0.2~MJ.

The Figures~\ref{HRR}-\ref{HRR2} have a shaded area that shows the 95\% confidence level.  The more narrow the shaded area, the better the repeatability of the fire.  This provides a quick means to assess the repeatability of a given source fire.  Using this measure the natural gas burner fires were the most repeatable, followed by the gasoline pan fires and then the polyurethane fires.

@@

In engineering terms, the ``flame height'' is typically discussed as the median flame height.  A position above the top of the fuel surface where the flame is observed to be present at least 50\% of the time.  As a starting point, the mean flame height for each fuel was estimated from a measurement of the uppermost continuous flame tip from the still photographic images.

@@

Another means of quantifying field data with Streams is the use of contour plots.  Examples of the time averaged contour plots are shown in Figures~\ref{Gas_and_Gasoline_Contours} and \ref{Foam_Contours} for the natural gas, gasoline, and polyurethane fueled fires.  Each contour line represents the percentage of time that the flame was at a given position. The region under examination, as shown as a black bounding rectangle, is 1400~mm high and 800~mm in width.  The natural gas fire was sampled for 29~s, resulting in 870~images. The area bounded by the contour with the value 1 represents the area that the flame was visible 100\% of the time sampled. This can be considered the persistent flame region as defined by McCaffrey~\cite{McCaffrey:1979}. The remaining contours are in the intermittent flame region, with the 0.5 contour representing the mean flame height. Outside of the final contour, value 0.0, no visible flame was recorded; hence this contour marks the definitive start of the buoyant plume region.

@@

Figure~\ref{Foam_Contours} shows two different contours for the polyurethane foam fueled experiment.  The results from polyurethane fueled fire characterization experiments have an increased amount of variability due to the nature of the ignition source and the fuel itself.  The polyurethane foam was readily ignited with a small pilot flame similar to that used to ignite the natural gas and the gasoline.  By comparison, a small flame ignited the gases flowing out of the top of the natural gas burner and the flames from the natural gas cover the entire burner surface within approximately one second.  A similar ignition sequence occurs with the gasoline pan fires. The polyurethane foam has a more complicated flame spread sequence that resulted in significant changes to the peak heat release rate and the time to peak heat release.  Once the polyurethane foam was ignited, it took approximately 60~s for the flame to spread across the top surface of the fuel.  As the flames spread horizontally across the fuel surface, the fire also burns down into the fuel.  In some cases the fire burned completely through the fuel in the center of the foam sample, resulting in a hollow flame at the time of the peak heat release rate.  An example of a contour plot, exhibiting the hollow or bifurcated flames is shown at right in Figure~\ref{Foam_Contours}. Refer back to Figure~\ref{flame_height_comp_FHPUF_1} to see photographic views of the bifurcated flames.  It seems that the movement of the flames as they flowed away from the fuel surface blurred out the bifurcation over time with the exception of the lower elevation areas that presented at the 90 and 100\% contours.  Other methods of ignition for the polyurethane foam were considered in order to have a more uniform ignition of the top surface of the foam.  Each method would have resulted in a larger release of energy from the ignition source or a starter fuel.   Based on a few scoping experiments these methods did not improve the repeatability.

Based on the field data developed using Streams average flame heights for each fuel were determined.  In Table~\ref{tab:Average_Flame_Heights}, each of the values represents the average flame heights, given in mm with 95\% confidence limits based on a Type A statistical analysis of the data, for each of the source fuels shown as the percentage of time the flame was visible at that given height.  For the natural gas fires, the averages are from ten 30~s segments, while the fire was monitored at approximately 80~kW.  In some cases more than one 30~s segment was sampled per heat release rate experiment.  Sixteen 30~s video segments, which included the peak heat release rate, were successfully analyzed from the seventeen gasoline heat release rate experiments.  One of the videos was unusable for analysis due to reflected light which could not be edited out of the images.  For the same reason, the polyurethane foam flame height averages are the result from eight of the ten heat release rate experiments.

\begin{table}
\centering
\begin{tabular}{|c|c|c|c|}
\hline
Visible         &   \multicolumn{3}{|c|}{Visible Flame Heights (mm)} \\ \cline{2-4}
Fraction (\%)   &       Natural Gas	            &   Gasoline	            & Polyurethane Foam \\ \hline \hline
10              &       980 $\pm$ 155 (16\%)   &	930 $\pm$ 130 (14\%) 	& 650 $\pm$ 220 (34\%)   \\
20              &   	890 $\pm$ 125 (14\%)   &	850 $\pm$ 120 (14\%) 	& 590 $\pm$ 190 (32\%)   \\
30              &   	820 $\pm$ 125 (15\%)   &	790 $\pm$ 110 (14\%) 	& 540 $\pm$ 180 (33\%)   \\
40              &   	760 $\pm$ 105 (14\%)   &	740 $\pm$ 95 (13\%) 	& 500 $\pm$ 165 (33\%)   \\
50              &   	705 $\pm$ 106 (15\%)   &	695 $\pm$ 95 (14\%) 	& 470 $\pm$ 160 (34\%)   \\
60              &   	650 $\pm$ 100 (15\%)   &	655 $\pm$ 85 (13\%) 	& 435 $\pm$ 135 (31\%)   \\
70              &   	590 $\pm$ 90 (15\%)    &	610 $\pm$ 90 (15\%) 	& 400 $\pm$ 125 (31\%)   \\
80              &   	530 $\pm$ 85 (16\%)    &	555 $\pm$ 90 (16\%) 	& 360 $\pm$ 110 (31\%)   \\
90              &   	440 $\pm$ 75 (17\%)    &	490 $\pm$ 95 (19\%) 	& 315 $\pm$ 110 (35\%)   \\
100	            &       220 $\pm$ 84 (38\%)    &	350 $\pm$ 110 (32\%) 	& 210 $\pm$ 125 (60\%)   \\
\hline
\end{tabular}
 \caption[Average flame heights for three fuels]{Average flame heights with 95\% confidence limits, for each of the source fuels shown as the percentage of time the flame was visible at that given height.}
 \label{tab:Average_Flame_Heights}
\end{table}
